\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-08 chapters/bttda.tex
Proofread

\section{Introduction}
\todo{reduce length of the introduction}

@@

Variants of HODA have been applied to BCI problems such as
ERP~\cite{Onishi2012,Higashi2016} and motor imagery (MI)~\cite{Liu2015,Cai2021}
decoding with positive results~\cite{Lotte2018}.
Recent work proposes adaptations such as suited objective
functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.
Discriminant tensor features have also been extracted
in the PARAFAC structure through manifold optimization~\cite{Froelich2018}.
However, it is not immediately obvious if either the Tucker or PARAFAC
structure are most suited to represent the neural data of interest for the BCI
paradigm and for decoding.

Recent research has shown that supervised decoders adopting a more flexible structure
can improve BCI performance.
Promising results have been achieved for regression tasks using
Higher Order Partial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022b}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaves room for improvement.
Instead of optimizing features directly for class separability, a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in these regression approaches are still more constrained
than what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low-rank common subspace
between the input and classification labels.
\textcite{Huang2020} propose a supervised approach for finding multiple discriminant
multilinear spectral filter terms and apply it to motor imagery BCI, but their
decomposition is also restricted in flexibility.

\todo{It's always tricky in the related work section of an ML/Mathematical
paper, but unless your reviewers are absolute experts of this specific field,
you should at least try to better describe the previous methods.}

@@

\subsection{Higher Order Discriminant Analysis}
Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products:
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
as visualized in \autoref{fig:HODA-backward}.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_backward.tikz.tex}
  \caption[A \acs{hoda} backward projection.]{%
    A visualization of the multilinear projection obtained by Higher Order
    Discriminant Analysis (HODA) applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		HODA finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}

@@

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$:
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices:
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which is solved through the HODA algorithm.
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per-mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor:
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem:
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices:
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of:
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
The iterative process halts after a fixed number of iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.
The full HODA procedure is summarized in \autoref{alg:HODA}.
\begin{algorithm}
  \caption[A \acs{hoda} backward solution.]{The \acs{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}

@@

The HODA projection in \autoref{eq:HODA-backward} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model, visualized in \autoref{fig:HODA-forward}, is given by:
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$, and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[A forward projection for \ac{hoda}.]{The forward projection for HODA.
    By calculating activation patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n)
  -\hat{\ten{X}}(n)\right)^2
  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmprs{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \autoref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

\subsection{Block-Term Tensor Discriminant Analysis}
After defining the forward model, we can construct our proposed block-term
tensor model.
Assuming the latent tensors $\ten{G}$
obtained by the backward projection of HODA do not achieve perfect
class separation, the error term $\ten{\mathbfcal{E}}$ in
\autoref{eq:HODA-forward} should still contain some discriminative
information, which can be exploited to improve classifier
performance.
Useful features can still be extracted from $\ten{\mathbfcal{E}} = \ten{X} -
\hat{\ten{X}}$, so it is further projected onto another core tensor
$\ten{G}^{(2)}$ (assuming $\ten{G}$ as $\ten{G}^{(1)}).$

We thus extend the HODA feature extraction scheme to Block-Term Tensor Discriminant Analysis
(BTTDA).
BTTDA finds multiple discriminative blocks, such that its forward
model adheres to the block-term tensor structure:
\begin{align}
	\ten{X} & = \sum_b^B\ten{G}^{(b)}\mmpr{\mat{A}^{(b)}} + \ten{\mathbfcal{E}}
	\label{eq:BTTDA-forward}
\end{align}
for $B$ extracted latent tensors $\ten{G}^{(b)}$ and residual error term
$\ten{\mathbfcal{E}}$.
\autoref{fig:BTTDA} further illustrates the BTTDA model.
\begin{figure*}[t]
	\centering
	\input{figures/bttda/BTTDA.tikz.tex}
  \caption[A forward model for \acs{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
		(BTTDA). BTTDA can extract more features
		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
		deflation scheme.
		The HODA backward projection is first applied. Next, the
		input data is reconstructed via the HODA forward model and the
		difference between the two is found.
		Finally, this process is repeated with this difference as input data, until a
		desired number of blocks $B$ has been found.}
	\label{fig:BTTDA}
\end{figure*}
The block-term structure of this model implies that it is a generalization of both
the Tucker-structured HODA and PARAFAC-structured discriminant feature
extraction.
If $B$ in \autoref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
HODA; if at each term $b$ the rank of the core tensor
$(r_1^{(b)}=r_2^{(b)}=\ldots=r_k^{(b)}=1)$, a PARAFAC structure is assumed.

Since BTTDA is specified above as a forward model, a backward procedure
is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in \autoref{alg:BTTDA}.
\begin{algorithm}
  \caption[A \acs{bttda} feature extraction.]{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \autoref{eq:HODA-backward} from the residual error term of the previous
block $\ten{\mathbfcal{E}}^{(b-1)}$:
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection:
\begin{equation}
  \ten{\mathbfcal{E}}^{(b)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.

@@

Combined, this results in hyperparameters $B$ for the number of blocks, the
threshold value for feature retention, and the ranks of the individual blocks.
While we can reasonably set the feature selection $p$-value threshold to $0.05$,
$B$ and the block ranks must be tuned to select a performant feature extraction
model.
While these hyperparameters can be set through cross-validation, this can be
computationally expensive.
To reduce the computational cost of model selection,
\autoref{alg:model-selection} proposes a heuristic model selection
algorithm that leverages cross-validation in a greedy way per block, to
iteratively find the optimal rank for the next block given the ranks of the
previous block.
The area under the receiver operating characteristic curve (ROC-AUC) for
classification of extracted feature vectors after feature selection is used as
a cross-validation score.
\begin{algorithm}
  \caption[Greedy model selection procedure.]{Greedy model selection.}
	\label{alg:model-selection}
	\input{chapters/bttda/alg_model_selection.tex}
\end{algorithm}
Finally, the series of blocks can be truncated to the point with
the highest validation score to determine $B$.

@@

\section{Experiments}
\subsection{Datasets and decoders}
We evaluated our proposed model in two offline EEG-based BCI decoding problems:
the event-related potential (ERP) and motor imagery (MI) paradigms using a
selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
Two ERP and two MI datasets were retained to reduce computational
demand.
Details about these datasets are found in \autoref{tab:moabb}.
For the ERP datasets, the task is to distinguish target from non-target ERPs,
while the MI datasets consist of distinguishing different imagined limb
movements.
Within-session classification performance was assessed using stratified 5-fold
cross-validation to calculate the area under the receiver operating
characteristic curve (ROC-AUC).
\begin{table*}[t]
	\centering
	\footnotesize
	\input{chapters/bttda/moabb.tex}
  \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
		subjects (\# Sub.), the number of EEG channels (\# Chan.), the number of
    classes (\# Classes), the number of trials or trials per class for ERP
    datasets (\# Trials), the epoch length (Epoch len.), the sampling
		frequency (S. freq.), the number of sessions per subject (\# Sess.) and the
		number of runs (\# Runs).
    Adapted from~\cite{Aristimunha2023}
    and~\cite{Chevallier2024}.}
	\label{tab:moabb}
\end{table*}

>>>>

\section{Introduction}

@@

Variants of HODA have been applied to BCI problems such as
ERP~\cite{Onishi2012,Higashi2016} and motor imagery (MI)~\cite{Liu2015,Cai2021}
decoding with positive results~\cite{Lotte2018}.
Recent work proposes adaptations such as suited objective
functions and regularization~\cite{JamshidiIdaji2017,Jorajuria2022,Aghili2023}.
Discriminant tensor features have also been extracted
in the PARAFAC structure through manifold optimization~\cite{Froelich2018}.
However, it is not immediately obvious if either the Tucker or PARAFAC
structure are most suited to represent the neural data of interest for the BCI
paradigm and for decoding.

Recent research has shown that supervised decoders adopting a more flexible structure
can improve BCI performance.
Promising results have been achieved for regression tasks using
Higher Order Partial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022a}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaves room for improvement.
Instead of optimizing features directly for class separability, a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in these regression approaches are still more constrained
than what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low-rank common subspace
between the input and classification labels.
\textcite{Huang2020} propose a supervised approach for finding multiple discriminant
multilinear spectral filter terms and apply it to motor imagery BCI, but their
decomposition is also restricted in flexibility.

@@

\subsection{Higher Order Discriminant Analysis}
Higher Order Discriminant Analysis (HODA)~\cite{Phan2010} is a
supervised tensor-based feature extraction technique.
For a set of $N$ tensors of order $K$
$\left\{\ten{X}(n)\in\mathbb{R}^{D_1\times D_2 \times \cdots \times
		D_K}\right\}_n^N$, HODA finds projection matrices $\mat{U_k}$ for each mode $k$
that project a given $\ten{X}$ to a latent tensor
$\ten{G}\in\mathbb{R}^{r_1\times r_2\times\cdots\times r_K}$, usually with lower
dimensionality $(r_1\leq D_1,r_2\leq D_2,\ldots,r_K\leq D_K)$ using
tensor-matrix mode products:
\begin{equation}
	\ten{G}  = \ten{X}\mmpr{\mat{U}}
	\label{eq:HODA-backward}
\end{equation}
as visualized in \cref{fig:HODA-backward}.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_backward.tikz.tex}
  \caption[A \acs{hoda} backward projection.]{%
    A visualization of the multilinear projection obtained by Higher Order
    Discriminant Analysis (HODA) applied to a third-order tensor
    sample $\ten{X}$ with shape $(D_1,D_2, D_3)$.
		HODA finds projection matrices $\mat{U}_k$ such that maximal
		discriminability between classes is achieved in the projected latent tensors
		$\ten{G}$ with reduced dimensionality $(r_1,r_2,r_3)$.}
	\label{fig:HODA-backward}
\end{figure}

@@

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$:
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices:
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which is solved through the HODA algorithm.
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per-mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor:
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix:
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem:
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices:
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of:
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
The iterative process halts after a fixed number of iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.
The full HODA procedure is summarized in \cref{alg:HODA}.
\begin{algorithm}
  \caption[A \acs{hoda} backward solution.]{The \acs{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}

@@

The HODA projection in \cref{eq:HODA-backward} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model, visualized in \cref{fig:HODA-forward}, is given by:
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$, and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[A forward projection for \ac{hoda}.]{The forward projection for HODA.
    By calculating activation patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\text{E}\left[
      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
    \right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \hat{\ten{X}}(n)\right)^2
  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \cref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

\subsection{Block-Term Tensor Discriminant Analysis}
After defining the forward model, we can construct our proposed block-term
tensor model.
Assuming the latent tensors $\ten{G}$
obtained by the backward projection of HODA do not achieve perfect
class separation, the error term $\ten{\mathbfcal{E}}$ in
\cref{eq:HODA-forward} should still contain some discriminative
information, which can be exploited to improve classifier
performance.
Useful features can still be extracted from $\ten{\mathbfcal{E}} = \ten{X} -
\hat{\ten{X}}$, so it is further projected onto another core tensor
$\ten{G}^{(2)}$ (assuming $\ten{G}$ as $\ten{G}^{(1)}).$

We thus extend the HODA feature extraction scheme to Block-Term Tensor Discriminant Analysis
(BTTDA).
BTTDA finds multiple discriminative blocks, such that its forward
model adheres to the block-term tensor structure:
\begin{align}
	\ten{X} & = \sum_b^B\ten{G}^{(b)}\mmpr{\mat{A}^{(b)}} + \ten{\mathbfcal{E}}
	\label{eq:BTTDA-forward}
\end{align}
for $B$ extracted latent tensors $\ten{G}^{(b)}$ and residual error term
$\ten{\mathbfcal{E}}$.
The BTTDA model is further illustrated by~\cref{fig:BTTDA}.
\begin{figure*}[t]
	\centering
	\input{figures/bttda/BTTDA.tikz.tex}
  \caption[A forward model for \acs{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
		(BTTDA). BTTDA can extract more features
		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
		deflation scheme.
		The HODA backward projection is first applied. Next, the
		input data is reconstructed via the HODA forward model and the
		difference between the two is found.
		Finally, this process is repeated with this difference as input data, until a
		desired number of blocks $B$ has been found.}
	\label{fig:BTTDA}
\end{figure*}
The block-term structure of this model implies that it is a generalization of both
the Tucker-structured HODA and PARAFAC-structured discriminant feature
extraction.
If $B$ in \cref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
HODA; if at each term $b$ the rank of the core tensor
$(r_1^{(b)}=r_2^{(b)}=\ldots=r_k^{(b)}=1)$, a PARAFAC structure is assumed.

Since BTTDA is specified above as a forward model, a backward procedure
is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in \cref{alg:BTTDA}.
\begin{algorithm}
  \caption[A \acs{bttda} feature extraction.]{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \cref{eq:HODA-backward} from the residual error term of the previous
block $\ten{\mathbfcal{E}}^{(b-1)}$:
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection:
\begin{equation}
  \ten{\mathbfcal{E}}^{(b)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.

@@

Combined, this results in hyperparameters $B$ for the number of blocks, the
threshold value for feature retention, and the ranks of the individual blocks.
While we can reasonably set the feature selection $p$-value threshold to $0.05$,
$B$ and the block ranks must be tuned to select a performant feature extraction
model.
While these hyperparameters can be set through cross-validation, this can be
computationally expensive.
To reduce the computational cost of model selection,
\cref{alg:model-selection} proposes a heuristic model selection
algorithm that leverages cross-validation in a greedy way per block, to
iteratively find the optimal rank for the next block given the ranks of the
previous block.
The area under the receiver operating characteristic curve (ROC-AUC) for
classification of extracted feature vectors after feature selection is used as
a cross-validation score.
\begin{algorithm}
  \caption[Greedy model selection procedure.]{Greedy model selection.}
	\label{alg:model-selection}
	\input{chapters/bttda/alg_model_selection.tex}
\end{algorithm}
Finally, the series of blocks can be truncated to the point with
the highest validation score to determine $B$.

@@

\section{Experiments}
\subsection{Datasets and decoders}
We evaluated our proposed model in two offline EEG-based BCI decoding problems:
the event-related potential (ERP) and motor imagery (MI) paradigms using a
selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
Two ERP and two MI datasets were retained to reduce computational
demand.
Details about these datasets are found in \cref{tab:moabb}.
For the ERP datasets, the task is to distinguish target from non-target ERPs,
while the MI datasets consist of distinguishing different imagined limb
movements.
Within-session classification performance was assessed using stratified 5-fold
cross-validation to calculate the area under the receiver operating
characteristic curve (ROC-AUC).
\begin{table*}[t]
	\centering
	\footnotesize
  \makebox[\textwidth][c]{
	  \input{chapters/bttda/moabb.tex}
  }
  \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
		subjects (\# Sub.), the number of EEG channels (\# Chan.), the number of
    classes (\# Classes), the number of trials or trials per class for ERP
    datasets (\# Trials), the epoch length (Epoch len.), the sampling
		frequency (S. freq.), the number of sessions per subject (\# Sess.) and the
		number of runs (\# Runs).
    Adapted from~\cite{Aristimunha2023}
    and~\cite{Chevallier2024}.}
	\label{tab:moabb}
\end{table*}

To use HODA as a decoder, it is paired with LDA to classify the
extracted feature (HODA+LDA), with hyperparameters $r_k$.
Similarly, we implemented BTTDA+LDA with the proposed BTTDA feature
extraction with hyperparameters $r_k^{(b)}$ for each block $b$ and the number of blocks
$1\leq B\leq16$.
Additionally, we also introduce PARAFACDA+LDA, which is the special case of
BTTDA+LDA where each $r_k^{(b)}=1$, with $B$ and the feature selection
threshold as only hyperparameters.
Hyperparameters were determined separately for each fold using nested
stratified 5-fold cross-validation, and, for BTTDA+LDA, in conjunction with the
greedy model selection algorithm in \cref{alg:model-selection}.
For HODA+LDA and the HODA blocks in BTTDA+LDA, we chose
$r=r_1=r_2=\ldots=r_K$ with possible values
$\textstyle{1,2,4,8,\ldots,\max_kD_k}$
to reduce computational cost.
Other HODA and BTTDA hyperparameters were set to
$B_\text{max}=10$, $\varepsilon=\num{1e-8}$ and $I_\text{max}=128$.
